\documentclass[11pt]{article}
\usepackage{foundation-backbone}

\begin{document}
\foundationtitle{Foundation as an Executable Application Backbone}{A technical guide for senior developers working across PostgreSQL, JavaScript, mobile, edge and native ecosystems}{Audience: senior developers and architects who may be new to Clojure, Lisp and Foundation}

\tableofcontents
\newpage

\section{Architectural intent}

Foundation Base is not primarily a web framework. It is a language and runtime toolkit for describing behaviour, emitting target-language code, running that code in target runtimes, and testing the agreement between those stages.

For application work, three areas form a useful backbone:

\begin{itemize}
  \item \product{postgres.core}: PostgreSQL definitions and generated database-side operations;
  \item \product{xt.substrate}: spaces, request/response transport, eventing and runtime context;
  \item \product{xt.db.node}: models, named views, local caches, SQL attachments, invalidation and remote refresh.
\end{itemize}

The intended outcome is a stable application contract that is executable before a production UI exists.

\section{Mental model for non-Lisp developers}

Lisp syntax makes code structure explicit with lists. A form such as:

\begin{lstlisting}[language=Lisp]
(model/view-refresh node "room/local" "orders" "list")
\end{lstlisting}

is equivalent in intent to a conventional function call. The additional value in Foundation is that forms can be inspected and transformed as data before they are emitted to JavaScript, SQL, Python or another target.

\begin{center}
\begin{tikzpicture}[node distance=11mm and 16mm]
  \node[fbwide] (form) {Source form\\data structure};
  \node[fbwide,right=of form] (prep) {Preprocessing\\macro expansion};
  \node[fbwide,right=of prep] (grammar) {Target grammar\\validated constructs};
  \node[fbwide,below=of prep] (emit) {Emitter output};
  \node[fbwide,left=of emit] (runtime) {Target runtime};
  \node[fbwide,right=of emit] (tests) {Emission and runtime tests};
  \draw[fbedge] (form) -- (prep);
  \draw[fbedge] (prep) -- (grammar);
  \draw[fbedge] (grammar) -- (emit);
  \draw[fbedge] (emit) -- (runtime);
  \draw[fbedge] (emit) -- (tests);
  \draw[fbedge] (runtime) -- (tests);
\end{tikzpicture}
\end{center}

This pipeline is why a portable module is more than a shared package. Its source form, transformations, target grammar, emitted syntax and runtime result are all part of the implementation.

\section{Backbone topology}

\subsection{Spaces and live views}

A substrate node owns application spaces. A space can represent a local client context, a remote server context, a test peer or another logical boundary. \product{xt.db.node} installs models into spaces. A model groups named views; each view has query input, status, value, dependencies and a query key.

\begin{center}
\begin{tikzpicture}[node distance=10mm and 18mm]
  \node[fbwide] (node) {substrate node};
  \node[fbwide,below left=of node] (local) {space: room/local};
  \node[fbwide,below right=of node] (server) {space: room/server};
  \node[fbbox,below left=of local] (list) {view: list};
  \node[fbbox,below right=of local] (detail) {view: detail};
  \node[fbwide,below=of server] (model) {server model};
  \draw[fbedge] (node) -- (local);
  \draw[fbedge] (node) -- (server);
  \draw[fbedge] (local) -- (list);
  \draw[fbedge] (local) -- (detail);
  \draw[fbedge] (server) -- (model);
\end{tikzpicture}
\end{center}

The application-facing shape is defined before storage. A view can begin in memory, then be backed by SQLite or delegated to a remote Postgres-backed space without changing the screen-level model.

\subsection{Local SQLite and remote PostgreSQL}

The repository walkthroughs exercise this topology directly:

\begin{center}
\begin{tikzpicture}[node distance=11mm and 20mm]
  \node[fbwide] (ui) {UI adapter};
  \node[fbwide,right=of ui] (lv) {local live view};
  \node[fbstorage,below=of lv] (sqlite) {SQLite};
  \node[fbwide,right=of lv] (transport) {substrate transport};
  \node[fbwide,right=of transport] (sv) {server view / RPC};
  \node[fbstorage,below=of sv] (pg) {Postgres};
  \draw[fbedge] (ui) -- node[above,scale=.8]{refresh / read} (lv);
  \draw[fbedge] (lv) -- (transport);
  \draw[fbedge] (transport) -- (sv);
  \draw[fbedge] (sv) -- (pg);
  \draw[fbedge] (lv) -- (sqlite);
  \draw[fbdashed,bend right=28] (sv.south) to node[below,scale=.8]{project result} (lv.south);
\end{tikzpicture}
\end{center}

This makes the local store an explicit materialised projection rather than an incidental framework cache.

\section{Role of \product{postgres.core}}

\product{postgres.core} is part of the target-language system rather than an ORM abstraction over SQL. PostgreSQL functions, graph operations and supporting definitions can be expressed in Foundation forms, emitted as SQL/PLpgSQL and executed in a PostgreSQL runtime.

A robust workflow treats database generation as a separate validation stage:

\begin{enumerate}
  \item inspect the source form and nearby grammar definitions;
  \item emit SQL and review the exact generated text;
  \item validate syntax and quoting against PostgreSQL;
  \item execute against a disposable database where the environment permits;
  \item invoke the operation through the same RPC path used by the application;
  \item verify the resulting local projection through \product{xt.db.node}.
\end{enumerate}

\begin{center}
\begin{tikzpicture}[node distance=9mm]
  \node[fbwide] (src) {Foundation Postgres form};
  \node[fbwide,below=of src] (sql) {Emitted SQL / PLpgSQL};
  \node[fbwide,below=of sql] (db) {Database execution};
  \node[fbwide,below=of db] (rpc) {RPC invocation};
  \node[fbwide,below=of rpc] (projection) {Local view projection};
  \draw[fbedge] (src) -- (sql);
  \draw[fbedge] (sql) -- (db);
  \draw[fbedge] (db) -- (rpc);
  \draw[fbedge] (rpc) -- (projection);
\end{tikzpicture}
\end{center}

\section{Testing strategy}

The highest-value tests are not isolated CRUD facts. They prove agreement across layers.

\subsection{Test pyramid for generated applications}

\begin{center}
\begin{tikzpicture}
  \filldraw[fill=FoundationLight,draw=FoundationBlue] (0,0) -- (12,0) -- (10.8,1.4) -- (1.2,1.4) -- cycle;
  \node at (6,.7) {grammar, transformation and emitter facts};
  \filldraw[fill=FoundationLight,draw=FoundationTeal] (1.2,1.6) -- (10.8,1.6) -- (9.5,3) -- (2.5,3) -- cycle;
  \node at (6,2.3) {target runtime and database execution};
  \filldraw[fill=FoundationLight,draw=FoundationGold] (2.5,3.2) -- (9.5,3.2) -- (8.2,4.6) -- (3.8,4.6) -- cycle;
  \node at (6,3.9) {RPC, sync and projection flows};
  \filldraw[fill=FoundationLight,draw=FoundationRed] (3.8,4.8) -- (8.2,4.8) -- (7.1,6.2) -- (4.9,6.2) -- cycle;
  \node[align=center] at (6,5.5) {thin UI\\acceptance};
\end{tikzpicture}
\end{center}

This ordering pushes correctness below the UI. UI tests then verify interaction and accessibility rather than re-proving every database rule.

\subsection{Representative flow test}

\begin{center}
\begin{tikzpicture}[node distance=12mm and 16mm]
  \node[fbbox] (test) {code.test fact};
  \node[fbwide,right=of test] (local) {local view refresh};
  \node[fbwide,right=of local] (remote) {remote dispatch};
  \node[fbstorage,below=of remote] (pg) {Postgres};
  \node[fbstorage,below=of local] (sqlite) {SQLite};
  \node[fbwide,left=of sqlite] (assert) {assert status, value, invalidation};
  \draw[fbedge] (test) -- (local);
  \draw[fbedge] (local) -- (remote);
  \draw[fbedge] (remote) -- (pg);
  \draw[fbedge] (remote) -- (local);
  \draw[fbedge] (local) -- (sqlite);
  \draw[fbedge] (sqlite) -- (assert);
\end{tikzpicture}
\end{center}

\section{AI-generated UI as an adapter}

Once the backbone is executable, UI generation can be treated as adapter generation. The AI receives:

\begin{itemize}
  \item model and view definitions;
  \item view inputs, statuses and result shapes;
  \item RPC names, arguments, errors and permissions;
  \item fixture data and passing flow tests;
  \item the design system and target framework constraints.
\end{itemize}

The generated client should not reproduce domain rules. It invokes operations, renders view state and adapts interaction to the target ecosystem.

\begin{center}
\begin{tikzpicture}[node distance=11mm and 15mm]
  \node[fbwide] (backbone) {Executable backbone};
  \node[fbwide,below=of backbone] (adapter) {Generated adapter boundary};
  \node[fbbox,below left=of adapter] (next) {Next.js};
  \node[fbbox,below=of adapter] (expo) {Expo / RN};
  \node[fbbox,below right=of adapter] (edge) {Bun / edge};
  \draw[fbedge] (backbone) -- (adapter);
  \draw[fbedge] (adapter) -- (next);
  \draw[fbedge] (adapter) -- (expo);
  \draw[fbedge] (adapter) -- (edge);
\end{tikzpicture}
\end{center}

\section{Comparison with established approaches}

\begin{tabularx}{\textwidth}{P{27mm}YYYY}
\toprule
\textbf{Approach} & \textbf{Contract} & \textbf{Local state} & \textbf{Generation} & \textbf{Main trade-off} \\
\midrule
REST + OpenAPI & HTTP operations and schemas & Client-framework responsibility & Excellent clients and server stubs & Weak model of view lifecycle and sync \\
GraphQL & Type graph and resolver operations & Normalised client cache, framework-specific & Strong schema tooling & Resolver/cache semantics are distributed \\
gRPC / Protobuf & Binary service contracts & Application responsibility & Excellent multi-language stubs & RPC-centric, not projection-centric \\
Prisma / ORM & Database model and generated client & Application responsibility & Strong DB client generation & Domain and sync behaviour remain elsewhere \\
Supabase/Firebase & Hosted schema, auth, realtime and SDK & Platform-managed or client SDK cache & Fast client delivery & Vendor-specific semantics and boundaries \\
CQRS/event sourcing & Commands, events and projections & Explicit projections & Often custom generators & Operational cost and eventual consistency complexity \\
Local-first frameworks & Replicated local model & First-class & Varies & Conflict semantics and backend integration can dominate \\
Foundation & Forms, grammars, spaces, views, RPC and runtime tests & Explicit local projection attached to a space & Multi-target emitters and portable modules & Requires Foundation literacy and cross-stage discipline \\
\bottomrule
\end{tabularx}

Foundation's differentiator is not that it alone can generate code or support local state. It is the ability to keep source forms, emitted target code, target runtimes and local/remote projection tests in one repository and one implementation workflow.

\section{Comparison with a Bun/Rust rewrite}

A Bun/Rust rewrite should be evaluated as an alternative decomposition, not as an automatic successor architecture.

\subsection{Typical Bun/Rust topology}

\begin{center}
\begin{tikzpicture}[node distance=11mm and 18mm]
  \node[fbwide] (client) {TypeScript clients};
  \node[fbwide,right=of client] (bun) {Bun API / edge layer};
  \node[fbwide,right=of bun] (rust) {Rust domain services};
  \node[fbstorage,right=of rust] (pg) {Postgres};
  \node[fbwide,below=of bun] (schema) {OpenAPI / Protobuf / shared schemas};
  \draw[fbedge] (client) -- (bun);
  \draw[fbedge] (bun) -- (rust);
  \draw[fbedge] (rust) -- (pg);
  \draw[fbdashed] (schema) -- (client);
  \draw[fbdashed] (schema) -- (bun);
  \draw[fbdashed] (schema) -- (rust);
\end{tikzpicture}
\end{center}

This architecture can be excellent when the system needs highly tuned native services, predictable memory use, low startup overhead or a TypeScript-native web/edge layer. It also makes the target implementation direct: Rust developers read Rust and Bun developers read TypeScript.

\subsection{Trade-off matrix}

\begin{tabularx}{\textwidth}{P{34mm}YY}
\toprule
\textbf{Dimension} & \textbf{Foundation} & \textbf{Bun/Rust rewrite} \\
\midrule
Primary abstraction & Portable source forms and target grammars & Services and packages in their implementation languages \\
Source of truth & Foundation module plus emitter/runtime tests & Protocol schemas plus Rust/TypeScript implementations \\
Runtime efficiency & Depends on emitted target and placement of work, especially PostgreSQL & Direct control; Rust is strong for CPU, memory and concurrency hot paths \\
Developer onboarding & Lisp syntax and Foundation concepts are unfamiliar to many & Bun/TypeScript is accessible; production Rust has a steep ownership/type learning curve \\
Cross-target consistency & High when portable modules and tests remain supported across targets & Must be enforced through schemas, generated clients and conformance suites \\
Debugging & Requires tracing source form through emitted code and runtime & Familiar target-native debugging in each service \\
Change surface & A model change may regenerate several targets & A contract change may require multiple service and client updates \\
Local/remote sync & Modelled directly through spaces, views and projections & Requires a chosen sync/local-first library or custom protocol \\
Best optimisation & Reduce semantic duplication and improve generation & Maximise target-native performance and operational isolation \\
Failure mode & Unsupported grammar constructs or emitter/runtime disagreement & Protocol drift, duplicated domain rules and service integration complexity \\
\bottomrule
\end{tabularx}

\subsection{Hybrid placement}

The strongest design may be hybrid:

\begin{center}
\begin{tikzpicture}[node distance=10mm and 17mm]
  \node[fbwide] (model) {Foundation contract and sync tests};
  \node[fbwide,below left=of model] (pg) {Generated Postgres RPC};
  \node[fbwide,below=of model] (bun) {Bun web / edge adapter};
  \node[fbwide,below right=of model] (rust) {Rust specialised service};
  \node[fbwide,below=20mm of model] (clients) {Web and mobile clients};
  \draw[fbedge] (model) -- (pg);
  \draw[fbedge] (model) -- (bun);
  \draw[fbedge] (model) -- (rust);
  \draw[fbedge] (pg) -- (clients);
  \draw[fbedge] (bun) -- (clients);
  \draw[fbedge] (rust) -- (clients);
\end{tikzpicture}
\end{center}

Use Rust for a measured hotspot or systems boundary, Bun for a TypeScript-native adapter, and Foundation for contract generation and end-to-end projection tests. This preserves the rewrite's performance benefits without discarding the tested backbone.

\section{Decision framework}

Prefer Foundation when:

\begin{itemize}
  \item the same domain behaviour must be projected across several target ecosystems;
  \item generated SQL and portable modules already express substantial business value;
  \item local SQLite/remote PostgreSQL behaviour is important;
  \item consistency and testable code generation are more valuable than target-native familiarity.
\end{itemize}

Prefer a direct Bun/Rust implementation when:

\begin{itemize}
  \item profiling demonstrates CPU, memory, latency or startup constraints that target-native code materially improves;
  \item the service boundary is narrow and stable;
  \item the organisation can sustain Rust and TypeScript implementations plus protocol conformance tests;
  \item local projection and cross-target generation are not central requirements.
\end{itemize}

Prefer a hybrid when a small number of native services provide clear value but the broader application still benefits from one tested contract.

\section{Implementation workflow}

\begin{enumerate}
  \item Read the relevant \product{xt.substrate}, \product{xt.db.node} and \product{postgres.core} source and matching \product{code.test} facts.
  \item Reproduce one vertical flow using the narrowest test namespace.
  \item Define the server model and RPC operation.
  \item Define local views and their remote space mapping.
  \item Attach Postgres to the server space and SQLite or cache storage to the local space.
  \item Assert emitted SQL separately from database execution.
  \item Assert refresh, invalidation and projection behaviour through the node runtime.
  \item Generate a thin UI adapter and verify it against fixtures and the live backbone.
  \item Profile before assigning work to Bun or Rust.
  \item Add contract/conformance tests for every specialised runtime boundary.
\end{enumerate}

\section{Conclusion}

Foundation offers a semantics-first route to cross-cutting applications. It is strongest where code generation, database-side RPC, portable modules and local/remote state must evolve together. Bun and Rust offer a runtime-first route with excellent target-native characteristics. A disciplined architecture chooses between them per boundary, based on measured constraints, while retaining executable contracts and end-to-end data movement tests as the common quality mechanism.

\end{document}
